### 2023-04-19 paper.Rmd
more work on problem formalization

where $t = 2$ gives the variance, $t = 3$ the skewness, and $t = 4$ the kurtosis [@joanes1998]. Thus, by changing $t$---the power of the deviation between mean and data points---we can use the k-plus logic to equalize any desired sample moment between groups. That is, we compute new variables for each original variable and for each distribution moment that should be considered in the anticlustering process. To formalize this idea, I now provide some notation to compute a generalized $\mathit{SSE}$ for k-plus anticlustering, which can include arbitrary distribution moments as part of the criterion. First, we define the $j$'th column ($j = 1, \ldots, P$) of the original data matrix $\mathbf{X}_{N \times P}$ as $x_{.j} =  (x_{1j}, \ldots, x_{Nj})$ where the column mean is $\overline{\mathbf{x}}_j$. Let $T$ be the number of distribution moments that should be equalized between $K$ groups. For each moment $t$ ($t = 1, \ldots, T$) and for each variable $x_{.j}$, a new *t-moment variable* $\hat{x}_{t.j} = (\hat{x}_{t1j}, \ldots, \hat{x}_{tNj})$ is defined where

@@

This computation results in an augmented data matrix $\mathbf{\hat{X}}_{N \times \mathit{TP}} = \{\hat{x}_{tij}\}_{N \times \mathit{TP}}$, i.e.

$$
  \mathbf{\hat{X}}_{N \times \mathit{TP}} =
  \left[ {\begin{array}{ccc|ccc|c|ccc}
    x_{11} & \cdots & x_{1P} & 
        (x_{11} - \overline{\mathbf{x}}_1)^2 & \cdots & (x_{1P} - \overline{\mathbf{x}}_P)^2 & 
        \cdots & 
        (x_{11} - \overline{\mathbf{x}}_1)^T & \cdots & (x_{1P} - \overline{\mathbf{x}}_P)^T \\
    x_{21} & \cdots & x_{2P} & 
        (x_{21} - \overline{\mathbf{x}}_1)^2 & \cdots & (x_{2P} - \overline{\mathbf{x}}_P)^2 & 
        \cdots & 
        (x_{21} - \overline{\mathbf{x}}_1)^T & \cdots & (x_{2P} - \overline{\mathbf{x}}_P)^T \\
    \vdots & \ddots & \vdots & \vdots & \ddots & \vdots & \cdots & \vdots & \ddots & \vdots\\
    x_{N1} & \cdots & x_{NP} & 
        (x_{N1} - \overline{\mathbf{x}}_1)^2 & \cdots & (x_{NP} - \overline{\mathbf{x}}_P)^2 & 
        \cdots & 
        (x_{N1} - \overline{\mathbf{x}}_1)^T & \cdots & (x_{NP} - \overline{\mathbf{x}}_P)^T \\
  \end{array} } \right]
$$

Note that when defining $\hat{x}_{tij}$, the condition $t = 1$ is separately included to keep the unchanged original data as the first $P$ columns in the data, which is used to equalize the the mean---the first moment---between groups. Using the data matrix $\mathbf{\hat{X}}_{N \times \mathit{TP}}$ and a vector of weights $\mathbf{w} = (w_1, \dots, w_T$), the generalized multicriterion for k-plus anticlustering $\mathit{SSE_{kplusGeneralized}}$ can be formalized:

@@

**TODO: include discussion of weight selection, as suggested by reviewer 2**

**argue why unweighted standardization k-plus anticlustering is good, as demanded by reviewer 2; also maybe include small scale simulation study**

Some simplifications can be arranged when ignoring the weights $w_1$ and $w_2$, e.g., by setting them both to 1. This simplification leads to the *unweighted sum approach* for k-plus anticlustering: When defining a new data matrix $\mathbf{Z}_{N \times 2P} = \{z_{ij}\}_{N \times 2P}$ as the augmentation of $\mathbf{X}_{N \times P}$ with $\mathbf{Y}_{N \times P}$, i.e.

$$
  \mathbf{Z}_{N \times 2P} =
  \left[ {\begin{array}{cccc|cccc}
    x_{11} & x_{12} & \cdots & x_{1P} & y_{11} & y_{12} & \cdots & y_{1P}\\
    x_{21} & x_{22} & \cdots & x_{2P} & y_{21} & y_{22} & \cdots & y_{2P}\\
    \vdots & \vdots & \ddots & \vdots & \vdots & \vdots & \ddots & \vdots\\
    x_{N1} & x_{N2} & \cdots & x_{NP} & y_{N1} & y_{N2} & \cdots & y_{NP}\\
  \end{array} } \right]
$$

and $z_{i.} = (x_{i1}, \ldots, x_{iP}, y_{i1}, \ldots, y_{iP})$, the bicriterion $\mathit{SSE_{kplus}}$ can be computed as

$$
\mathit{SSE_{kplus}} =
  \sum\limits_{j=1}^{2 P} 
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (z_{ij} - \overline{z}_j ^{(k)})^2.
$$

This term again has the same form as $\mathit{SSE}$ but is computed on different data. Hence, computing the k-plus criterion can be done by first computing the squared distances $y_{ij}$, appending them as new columns to the original data matrix, and then computing $\mathit{SSE}$ on the augmented data. Using the unweighted sum approach, k-plus anticlustering therefore leaves the original k-means criterion unchanged.

Note that a problem may arise if the weights $w_1$ and $w_2$ are ignored when computing $\mathit{SSE_{kplus}}$. Because the computation of the new attribute values $y_{ij}$ involves squaring, their range of values is expected to be very different from the raw values $x_{ij}$ (e.g., see Figure 1). This imbalance can strongly and unpredictably influence the relative weight of $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ when computing $\mathit{SSE_{kplus}}$. To ensure an equal weight of both criteria, all attributes ($x_{ij}$ and $y_{ij}$) can be standardized before computing $\mathit{SSE_{kplus}}$, e.g. via $z$-standardization.

>>>>

where $t = 2$ gives the variance, $t = 3$ the skewness, and $t = 4$ the kurtosis [@joanes1998]. Thus, by changing $t$---the power of the deviation between mean and data points---we can use the k-plus logic to equalize any desired sample moment between groups. That is, we compute new variables for each original variable and for each distribution moment that should be considered in the anticlustering process.

To formalize the computation of the k-plus criterion for an arbitraty number of distribution moments, some new notation has to be provided. First, we define the $j$'th column ($j = 1, \ldots, P$) of the original data matrix $\mathbf{X}_{N \times P}$ as $x_{.j} =  (x_{1j}, \ldots, x_{Nj})$ where the column mean is $\overline{\mathbf{x}}_j$. Let $T$ be the number of distribution moments that should be equalized between $K$ groups. For each moment $t$ ($t = 1, \ldots, T$) and for each variable $x_{.j}$, a new *t-moment variable* $\hat{x}_{t.j} = (\hat{x}_{t1j}, \ldots, \hat{x}_{tNj})$ is defined where

@@

This computation results in a three dimensional tensor $\mathbf{\hat{X}}_{T \times N \times P}$ = $\{\hat{x}_{tij}\}_{T \times N \times P}$ where---in comparison to the original data matrix $\mathbf{X}_{N \times P}$---the additional dimension is given by the $T$ distribution moments that are considered in the anticlustering process. When defining $\hat{x}_{tij}$, the condition $t = 1$ is separately included to keep the unchanged original data as first entry in the "moment dimension" in $\mathbf{\hat{X}}_{T \times N \times P}$, which is used to equalize the the mean---the first moment---between groups.

Using the tensor  $\mathbf{\hat{X}}_{T \times N \times P}$ and a vector of weights $\mathbf{w} = (w_1, \dots, w_T$), the generalized multicriterion for k-plus anticlustering $\mathit{SSE_{kplusGeneralized}}$ can be formalized:

@@

Some simplifications in the computation of $\mathit{SSE_{kplusGeneralized}}$ can be arranged when ignoring the moment weights $\mathbf{w}$, e.g. by setting $w_t = 1$ ($\forall t = 1, \dots, T$). This simplification leads to the *unweighted sum approach* for k-plus anticlustering. In this case, the additional dimension in $\mathbf{\hat{X}}_{T \times N \times P}$, representing distribution moments can be dropped; instead, the tensor is flattened into a two dimensional matrix, where the variables representing the additional distribution moments are appended to the original data matrix $\mathbf{X}_{N \times P}$, resulting in the augmented matrix $\mathbf{\hat{X}'}_{N \times \mathit{TP}}$:

$$
  \mathbf{\hat{X}'}_{N \times \mathit{TP}} =
  \left[ {\begin{array}{ccc|ccc|c|ccc}
    x_{11} & \cdots & x_{1P} & 
        (x_{11} - \overline{\mathbf{x}}_1)^2 & \cdots & (x_{1P} - \overline{\mathbf{x}}_P)^2 & 
        \cdots & 
        (x_{11} - \overline{\mathbf{x}}_1)^T & \cdots & (x_{1P} - \overline{\mathbf{x}}_P)^T \\
    x_{21} & \cdots & x_{2P} & 
        (x_{21} - \overline{\mathbf{x}}_1)^2 & \cdots & (x_{2P} - \overline{\mathbf{x}}_P)^2 & 
        \cdots & 
        (x_{21} - \overline{\mathbf{x}}_1)^T & \cdots & (x_{2P} - \overline{\mathbf{x}}_P)^T \\
    \vdots & \ddots & \vdots & \vdots & \ddots & \vdots & \cdots & \vdots & \ddots & \vdots\\
    x_{N1} & \cdots & x_{NP} & 
        (x_{N1} - \overline{\mathbf{x}}_1)^2 & \cdots & (x_{NP} - \overline{\mathbf{x}}_P)^2 & 
        \cdots & 
        (x_{N1} - \overline{\mathbf{x}}_1)^T & \cdots & (x_{NP} - \overline{\mathbf{x}}_P)^T \\
  \end{array} } \right]
$$

When defining the row vectors of $\mathbf{\hat{X}'}_{N \times \mathit{TP}}$ with the notation $\hat{x}'_{i.} = (\hat{x}'_{i1}, \ldots, \hat{x}'_{i(TP)})$ for each $i = 1, \ldots, N$, the generalized criterion $\mathit{SSE_{kplusGeneralized}}$ can be computed as

$$
\mathit{SSE_{kplusGeneralized}} =
  \sum\limits_{j=1}^{\mathit{TP}}
  \sum\limits_{k=1}^{K} 
  \sum\limits_{i \in C_k}
  (\hat{x}'_{ij} - \overline{\hat{x}'}_j ^{(k)})^2.
$$

Thus, when setting all the weights for all moments to 1, the k-plus criterion $\mathit{SSE_{kplusGeneralized}}$ again has the same form as the standard k-means criterion $\mathit{SSE}$, but it is computed on different data. Here, computing the k-plus criterion is done by first computing the additional $t$-moment variables, appending them as new columns to the original data matrix, and then computing $\mathit{SSE}$ on the augmented data. Using the unweighted sum approach, k-plus anticlustering therefore leaves the original k-means criterion unchanged.

Note that a problem may arise if the weights $\mathbf{w}$ are ignored when computing. Because the computation of the new $t$-moment variables involves squaring, their range of values is expected to be very different from the raw values $x_{ij}$ (e.g., see Figure 1). This imbalance can strongly and unpredictably influence the relative weight of $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ when computing $\mathit{SSE_{kplus}}$. To ensure an equal weight of both criteria, all attributes---i.e., the original data as well as the new $t$-moment varibles---can be standardized before computing $\mathit{SSE_{kplusGeneralized}}$, e.g. via $z$-standardization.
